---
title: "Reproducible Research - Peer Assessments 1"
author: "Saul Lugo"
date: "September 19, 2015"
output: html_document
---

First load "knitr" and set the global options for knitr

```{r set_options}
require(knitr)
opts_chunk$set(echo=TRUE)
```

The data will be loaded from the zip file provided in the course, and the resulting csv will be stored inside a directory named "data" in your working directory.

```{r load_data}
if(!file.exists("./data/activity.csv")){
        if(!file.exists("./data/")) dir.create("data")
        if(!file.exists("activity.zip")) stop("The data file does not exists in the directory, you must download the file activity.zip")
        unzip("activity.zip",exdir="data")
}
#Reading the dataset
data <- read.csv("./data/activity.csv")
```

##First Question: What is the mean total number of steps taken per day?

1) Calculating the total number of steps taken per day.In this part of the assigment I am ignoring the NAs values in the dataset.

```{r steps}
steps_per_day <- tapply(data$steps,data$date,sum)
```

2) Making a histogram of steps_per_day. This will make a graphic representation of how frequent a certain number of steps was taken by the individual in a day during the measurement period.

```{r steps_hist}
hist(steps_per_day,main="Number of Steps per Day",xlab="Number of Steps",breaks = 10,freq=TRUE,col="blue")
```

3) Calculating the mean and median of the total number of steps taken by day

```{r steps_mean_median}
#The mean of the total number of steps taken each day is:
mean_steps <- mean(steps_per_day[!is.na(steps_per_day)])
mean_steps
#The median of total number of steps taken each day is:
median_steps <- median(steps_per_day[!is.na(steps_per_day)])
median_steps
```

