---
title: "Reproducible Research - Peer Assessments 1"
author: "Saul Lugo"
date: "September 19, 2015"
output: html_document
---

First load "knitr" and set the global options for knitr

```{r set_options}
require(knitr)
opts_chunk$set(echo=TRUE)
```

The data will be loaded from the zip file provided in the course, and the resulting csv will be stored inside a directory named "data" in your working directory.

```{r load_data}
if(!file.exists("./data/activity.csv")){
        if(!file.exists("./data/")) dir.create("data")
        if(!file.exists("activity.zip")) stop("The data file does not exists in the directory, you must download the file activity.zip")
        unzip("activity.zip",exdir="data")
}
#Reading the dataset
data <- read.csv("./data/activity.csv")
```

##Question 1: What is the mean total number of steps taken per day?

1) Calculating the total number of steps taken per day.In this part of the assigment I am ignoring the NAs values in the dataset.

```{r steps}
steps_per_day <- tapply(data$steps,data$date,sum)
```

2) Making a histogram of steps_per_day. This will make a graphic representation of how frequent a certain number of steps was taken by the individual in a day during the measurement period.

```{r steps_hist}
hist(steps_per_day,main="Number of Steps per Day",xlab="Number of Steps",breaks = 10,freq=TRUE,col="blue")
```

3) Calculating the mean and median of the total number of steps taken by day

```{r steps_mean_median}
#The mean of the total number of steps taken each day is:
mean_steps <- mean(steps_per_day[!is.na(steps_per_day)])
mean_steps
#The median of total number of steps taken each day is:
median_steps <- median(steps_per_day[!is.na(steps_per_day)])
median_steps
```

##Question 2: What is the average daily activity pattern?

1) Making a time series plot of 5-min interval (x-axis) and the average number of steps taken, averaged across all days (y-axis)

```{r activity_pattern}
mean_steps_per_interval <- tapply(data$steps,data$interval,mean,na.rm=TRUE)
plot(mean_steps_per_interval,type="l",col="blue",main="Average steps taken per 5-min interval")
```

2) Which 5-min interval, on average across all the days in the dataset, contains the maximum number of steps?
To answer this question I used the function "which.max"
```{r max_number_steps}
x <- which.max(mean_steps_per_interval)
"The 5-min that contains, in average, the maximum number of steps is the interval number:"
x[[1]]
```

##Question 3: Imputing missiong values

Note that there are a number of days/intervals where there are missing values (coded as NA). The presence of missing days may introduce bias into some calculations or summaries of the data.

1) Calculate and report the total number of missing values in the dataset (i.e.the total number of rows with NAs)
```{r missing_values}
n_missing_values <- length(data[is.na(data$steps),]$steps)
"Number of row with missing values"
n_missing_values
```

2) Devise a strategy for filling in all of the missing values in the dataset. The strategy does not need to be sophisticated. For example, you could use the mean/median for that day, or the mean for that 5-minute interval, etc.

```{r filling_missing_values}
require(dplyr)
stepsMissing <- function(d){
        vec <- vector("numeric",length=length(d$steps))
        
        for(i in seq_along(vec)){
                if(is.na(d$steps[i])){
                        m <- tapply(d$steps,d$interval==d$interval[i],mean,na.rm=TRUE)
                        vec[i] <- m[[2]]
                }else{
                        vec[i] <- d$steps[i]
                }
        }
        vec
}

fixed_data <- mutate(data, steps = stepsMissing(data))
head(fixed_data)
```

